Lattice Boltzmann study of mass transfer for Bretherton/Taylor bubble train flow

### 2012-02-21 Text/introduction.tex
Finished another iteration.

\begin{document}
\maketitle
\begin{abstract}
This work presents a thorough procedure of how to determine the volumetric mass transfer
coefficient
in the context of lattice Boltzmann simulations for Bretherton/Taylor bubble train motion. The
novelty of this work lies in the establishing
a thorough procedure of how to measure the volumetric mass transfer coefficient addressing the
following challenges: the flow pattern  depends on the capillary number $Ca$ regime, i.e. for
$Ca<0.7$ the flow has a vortex in front of the
bubble \cite{giavedoni-numerical}, but for the larger capillary number it hasn't;  the mass sources are comparable of the
representative unit cell size (inhomogenous situation is different from the homogenous well-mixed
media \cite{jos-mass}); the connection between time and space domains for moving bubbles; the
definition of the mass transfer coefficient which is defined through the characteristic
concentration (inlet/outlet concentration, average concentration). The following boundary conditions
were examined: periodic boundary conditions, open boundaries, and a few unit cells open boundary
simulations. It was shown that the time dependent average concentration to be a characteristic
concentration produce robust results, though a bit overpredicted. The inlet/outlet flux averaged
concentration to be a characteristic concentration produce unreliable results. In comparison, a few
unit cells open boundaries conditions resemble physical situation most and are harder to conduct
but provide the volumetric mass transfer coefficient based only on the spatial information for the
capillary number $Ca>0.7$. As well, a few unit cells simulations can provide more information about
mixing in the liquid slug in comparision with the periodic boundary conditions. We show
that all presented in literature strategies are extreme limits of one equation,
Eq. {\color{red} give equation number here}. Finally, the correlation for different Peclet numbers
is compared with analytical predictions presented by \citet{vanbaten-circular} and shows a good
comparison. The work is of use to people performing mass transfer numerical simulations for bubble
train flow in microchannels and within the
lattice Boltzmann context.
\end{abstract}

\section{Introduction}
The monolith reactors during last years are getting more attention as alternative to slurry
reactors and trickle bed reactors \cite{kreutzer-overview,bercic-mass}. The bubble flow we focus in
this work, so-called Bretherton-Taylor flow \cite{bretherton,taylor}, describes the flow through
a liquid medium of equal sized, long in comparison with the channel height bubbles, see
Fig. \ref{fig:benchmark:hydro}. This flow regime
is usually characterized by surface tension dominance over inertia and viscous effects and by
comparatively small gas flow velocities \cite{yue-mass}. Due to surface tension dominance and small
scales of microchannels, bubble train flow exhibit advantageous properties which can not be achieved
in macroscopic counterparts: such as thin films \cite{bretherton} can remarkably enhance mass
transfer in the film from a gas or from walls; the plug flow nature that separates liquid slugs from each
other allowing to perform chemical reactions in slugs \cite{kreutzer-overview} or be an indicator of
the liquid velocity \cite{taylor}; the flow pattern depending on the capillary number $Ca$
\cite{heil-bretherton,heil-threedim}.  Overall, nowadays one can find a large number of applications
of Bretherton-Taylor bubble train flow: continuous flow analyzers, lung openings, chemical reactions
as hydrogenation of nitroaromatics, 2-ethyl-hexenal, Fischer-Tropsch synthesis, etc. One can find a
lot of different application references in extensive reviews \cite{kreutzer-overview,
gupta-review,yue-mass}.

@@

There are different methods to estimate the mass transfer coefficient $k_L$. We first examine the
theoretical definitions of the mass transfer in case of point mass sources.
\subsection{Point mass sources}
In what follows we will present three approaches to calculate point mass transfer
coefficients:
\begin{enumerate}
\item
Let us look at the infinitesimal small domain of the volume $A \Delta x$ not
moving and with the point mass source.  Then the concentration difference can be found as $\Delta C =
\cstar -
C(t)$, where $\cstar$ is the imposed point source concentration, $C(t)$ is
the time dependent concentration, which do not depend on the location due to homogeneity. Therefore, one can write the time dependent ODE for the
concentration in the domain:
\beq
\dot{m}= A \Delta x \frac{\mathrm{d}C}{\mathrm{d} t} = k_L P (\cstar-C(t)), 
\feq
with the initial condition $C(0)=0$
The solution can be found by solving ODE:
\beq
C(t)= \cstar (1-\exp(-\vol t )), 
\feq
where $\vol$ is the volumetric mass transfer coefficient defined as:
\beq
\vol=k_L \frac{P}{V},
\feq
where $P$ is the source surface, $V$ is the unit cell volume.
\item
The previous consideration is able to predict mass transfer happening in moving with the velocity
$U$ liquid, see Fig. \ref{fig:moving:frame}. 
\begin{figure}[htb!]
\includegraphics[width=\textwidth]{Figures/mass_transfer.eps}
\caption{The mass transfer in the moving liquid. \label{fig:moving:frame}}
\end{figure}

If one can assume that the mass point sources are distributed in the whole medium, then the mass accumulated in the unit volume $V=A \Delta x$ can be calculated as the difference of 
mass fluxes entering and leaving domain $U \bigl(C(x+\Delta x)-C(x)\bigr)$. We assume that the
source is a point source with the effective perimeter (area) $P$. The accumulated mass should be proportional
to the mass transfer coefficient:
\beq
U \bigl(C(x+\Delta x)-C(x)\bigr)=k_L P \bigl(\cstar-C(x)\bigr), 
\feq 
giving the same equation but only in the coordinate domain:
\beq
%\label{concentration:domain:coordinate}
C(x)= \cstar \Bigl(1-\exp\bigl(-k_L a \frac{x}{U} \bigr)\Bigr).
\label{main:mass:transfer:expression} 
\feq

@@

\subsection{Bubble train}
In the application to the bubble train flow, if one can think of one bubble to be a point source,
then all considerations above are valid. For example, the expression
(\ref{main:mass:transfer:expression}) was used in the experiments by
\citet{bercic-mass}. However, one should be accurate with the definition of velocities because two
different phases co-exist in the bubble train flow. Usually, one can take the velocity $U$ to be as
a bulk velocity or $U=\ugas+\uliq$, where $\ugas$ and $\uliq$ are liquid and gas
superficial velocities correspondingly.

Though the experimental measurements are straightforward as measuring concentrations at different
locations and calculating the volumetric mass transfer coefficient through the logarithmic
function. However, if one wants to analytically estimate the mass transfer coefficients, the
situation is much more complicated because of two phases presence and complicated bubble
geometry. Phases can have different mixing patterns
from film to slug and from slug to film. As it was mentioned before depending on the capillary
number the mixing velocity pattern is different. Analytical approaches
\cite{irandoust,vanbaten-circular} assume that the
contributions from film and bubble caps can be calculated separately. However, one can see that
such 
an assumption for bubble caps (slug well mixed and the concentration is uniformly distributed over
the slug) usually overpredicts the mass transfer \cite{irandoust}. This happens since some tracer
concentration from film is mixed with the slug and changes the overall concentration in the slug.
Thus, all estimations for the analytical mass as well do not account mutual mass transfer from
neighbouring bubbles.

Another issue is that if the slug is long enough then the film saturates with the tracer
concentration and its influence on the mass transfer can be negligible \cite{vanbaten-circular}.
Mixing patterns of the film and the liquid slugs are of great importance for the analytical
estimation of the mass transfer \cite{yue-mass}. However, the assumptions usually taken for the
mass transfer calculation are small capillary number and certain mixing patterns which help to
estimate the mass transfer using the penetration theory of \citet{higbie}.

The numerical approach can take into the account the complicated mixing patterns and geometries.
However, there is a  challenge to restore the continous approach as it is seen in experiments. Next
section gives more details about numerical simulations.

\subsection{Numerical simulations}
\label{section:cases}
To restore the continuous picture one needs to perform a number of bubbles simulations. As we saw
before there are two approaches towards it - either to simulate
the bubble train and then to measure concentration along the pipe, Eq.
\ref{main:mass:transfer:expression}, or to transfer to the reference frame moving with the bulk
velocity $U$ and conduct same measurements. However, both methods do require a tracking of
moving bubbles which is complicated from the numerical point of view. Therefore, one needs to come
up with the simple smaller domain calculations of the mass transfer coefficient, which mimic the
continuous picture with infinite numbers of the bubbles closely.

@@

We perform simulations in the frame moving with the bubble (velocity
$\ububble$), that the bubble position stands steady. The bubble velocity $\ububble$ is
different from the bulk velocity $\ugas+\uliq$. In this case we need to perform a coordinate $x$
variable change:
\beqal
\label{theor:average:concentration:time}
&x(t)=\ububble t\\
&C(x)=\cstar \Bigl(1-\exp\bigl(-\vol \frac{x}{\ugas+\uliq}\bigr)\Bigr)\\
&C(t)=\cstar \Bigl(1-\exp\bigl(-\vol\, t \frac{\ububble}{\ugas+\uliq}\bigr)\Bigr).
\feqal
Therefore, one can obtain the volumetric mass transfer coefficient through the concentration:
\beqal
\label{theor:one:concentration:time}
&\vol\, t \frac{\ububble}{\ugas+\uliq}=\ln \frac{\cstar}{\cstar-C(t)}\\
&\volnondim=\frac{\lunit}{\ububble t}\ln \frac{\cstar}{\cstar-C(t)},
\feqal
where the parameter $\vol \frac{\lunit}{\ugas+\uliq}$ is non-dimensional. One can also measure the
volumetric mass transfer coefficient from concentrations given at times $t_1$ and $t_2$:
\beq
\label{theor:continuous:mass:transfer}
\volnondim=\frac{\lunit}{\ububble
(t_2-t_1)}\ln\frac{C^{*}-C(t_1)}{C^{*}-C(t_2)}.
\feq
Expressions (\ref{theor:average:concentration:time}) and (\ref{theor:continuous:mass:transfer}) are
the cornerstones of the present work. The main questions arisen in the present work are to define
$C(t)$ and  the boundary conditions to produce a proper volumetric mass transfer coefficient
$\vol$.

Three possible scenarios of conducting numerical simulations are possible: 
\begin{enumerate}
\item As far as the hydrodynamic flow is periodic, one can make an assumption of the same boundary
conditions applied to the tracer concentration. Therefore, only one unit cell is simulated with
periodic boundary conditions, see Fig. \ref{fig:benchmark}. As well, no tracer leaves a
domain. The situation resembles a situation for the plug flow, where no tracer can leave a unit
domain. We expect such a situation happening for small Capillary numbers $Ca<0.1$, where the film
is thin and saturates fast with the tracer. Though easier to implement, it rises certain critisism
towards the inlet concentration to be equal to the outlet concentration. In the continuous
formulation there is the concentration difference between inlet and outlet involving the volumetric
mass transfer coefficient:
\beq
\volnondim=\ln\frac{\cstar-\cinlet}{\cstar-\coutlet}.
\feq
The concentration $C(t)$ required for calculation the volumetric mass transfer coefficient, Eq.
\ref{theor:one:concentration:time}, can be taken as the average concentration in the domain:
\beq
\label{average:concentration:definition}
C(t)=\frac{\int_{liquid}{C \mathrm{d}V}}{\int{\mathrm{d}V}}.
\feq
\begin{figure}[htb!]
\includegraphics[width=\textwidth]{Figures/benchmark_middle.eps}\\
\caption{The two-dimensional benchmarks for the  the mass transfer
coefficient (bottom) for the bubble located near the entrance (top) and at the middle of the domain
(bottom). \label{fig:benchmark}}
\end{figure}
\item The second approach is the same as the first one but the concentration is taken as in
\cite{vanbaten-circular}. The characteristic concentration is taken as the inlet/outlet
concentration:
\beqal
&\cinlet=\frac{\int{U(y) C(0,y,t) \mathrm{d}y}}{\int{U(0,y) \mathrm{d}y}}\\
&\coutlet=\frac{\int{U(y) C(\lunit,y,t) \mathrm{d}y}}{\int{U(\lunit,y) \mathrm{d}y}}\\
&\cinlet=\coutlet.
\feqal
Therefore the assumptions  of this approach is that the difference between inlet and outlet is not
considerably large and the tracer is well mixed in the slug.

\item
The approach of \citet{vanbaten-circular} is taken. Periodic boundary conditions were used. By the
definition the mass transfer coefficient is
proportional to the gain of the mass in the system divided by the concentration difference
multiplied on the surface area:
\beq
\vol=\frac{\dot{m}}{P \Delta C}\frac{P}{V}=\frac{\dot{m}}{V (\cstar-C(t))},
\feq
where mass difference in the domain can be calculated as:
\beq
\dot{m}=\frac{m_2-m_1}{t_2-t_1}=\frac{\int_{liq}{C(t_2)\mathrm{d}V}-\int_{liq}{C(t_1)\mathrm{d}V}}{
t_2-t_1 } .
\feq
In the approach of \citet{vanbaten-circular} the inlet and outlet concentrations were taken as the
characteristic concentration $C(t)$.

%\item 
%The approach of the opened boundaries was examined, see Fig. \ref{fig:benchmark:jos}. At the inlet
%and outlet boundaries the diffusion flux was imposed to be zero, $\frac{\partial C}{\partial x}=0$.
%The characteristic concentration was taken as the average concentration in the whole domain.
%\begin{figure}
%\includegraphics[width=\textwidth]{Figures/benchmark_jos.eps}
%\caption{The numerical mass transfer benchmark with open boundaries. The mass transfer is
%calculated according to Eq. \ref{theor:average:concentration:time} with the characteristic
%concentration accoring to Eq. \ref{average:concentration:definition}. \label{fig:benchmark:jos}}
%\end{figure}

\item 
The last approach to be examined in this paper is the approach of the simulation of a lot of unit
volumes, see Fig. \ref{fig:benchmark:alot}. This situation corresponds to the simulation of the
bubble train head, after injecting to the pipe and travelling along the pipe. The argument here is
that behind a certain number of the bubble train head, the influence of the boundaries will be
diminished. Thus, the average concentration in time will be governed by Eq.
\ref{theor:continuous:mass:transfer}. This simulation statement is of most correspondence to
experiments and should give right answer. However, one of the
disadvantages is to simulate a certain number of unit cells, thus increasing a computational
complexity.
\begin{figure}[htb!]
\includegraphics[width=\textwidth]{Figures/benchmark_alot.eps}\\
\caption{Benchmark for a lot of units. \label{fig:benchmark:alot}}
\end{figure}
\end{enumerate}

Before all the approaches will be examined we will perform benchmarks of parts of geometries that
form a problem.

\section{Validation}
As for analytical expressions the problem of the mass transfer can be decomposed to the following
parts: the mass transfer from two hemispheres and the mass transfer from the film. We will examine
these problems with the help of the lattice Boltzmann method and compare them against analytical
solutions. The next sections will present the lattice Boltzmann method and a few benchmarks.

\subsection{TRT $D2Q9$ model}
The lattice Boltzmann equation (LBE) operates on a rectangular grid representing the
physical domain. It utilizes
probability distribution functions (also known as particle populations)
containing information about
macroscopic variables, such as fluid density and momentum. LBE consists of
two parts: a local collision step, and a propagation step which transports
information from one node to another along some 
directions specified by the discrete velocity set.
The LBE is typically implemented as follows:
\begin{equation}
\label{standard:implementation}
\begin{aligned}
&f_i^{*}(\bm{x},t)= f_i(\bm{x},t)-\omega \bigl(f_i(\bm{x},t)-f_i^{eq}(\bm{x},t)\bigr),&&\text{
collision step}\\
&f_i(\bm{x}+\bm{c_i},t+1)=f_i^{*}(\bm{x},t),&&\text{ propagation step}, 
\end{aligned}
\end{equation}
where $f_i$ is the probability distribution function in the direction $\bm{c_i}$,
 $f_i^{eq}$ is the equilibrium probability distribution function, $\omega$ is the
relaxation parameter. Collision operator, $-\omega \bigl(f_i- f_i^{eq}\bigr)$ is so-called BGK
collision operator \cite{bgk}. However, the approach to be used here is the TRT
(two-relaxation-times) collision operator \cite{ginzburg-main,ginzburg-saturated-flow}. In
comparison with the widely used BGK collision operator, TRT collision operator have better accuracy
for diffusion and convection fluxes, larger range of parameters where the LBM scheme is stable.

>>>>

\begin{document}
\maketitle
\begin{abstract}
This work presents a thorough procedure of how to determine the volumetric mass transfer
coefficient
in the context of lattice Boltzmann simulations for Bretherton/Taylor bubble train motion. The
novelty of this work lies in the establishing
a thorough procedure of how to measure the volumetric mass transfer coefficient addressing the
following challenges: the flow pattern  depends on the capillary number $Ca$ regime, i.e. for
$Ca<0.7$ the flow has a vortex in front of the
bubble \cite{giavedoni-numerical}, but for the larger capillary number it hasn't;  the mass sources are comparable of the
representative unit cell size (inhomogenous situation is different from the homogenous well-mixed
media \cite{jos-mass}); the connection between time and space domains for moving bubbles; the
definition of the mass transfer coefficient which is defined through the characteristic
concentration (inlet/outlet concentration, average concentration). The following boundary conditions
were examined: periodic boundary conditions, open boundaries, and a few unit cells open boundary
simulations. It was shown that the time dependent average concentration to be a characteristic
concentration produce robust results, though a bit overpredicted. The inlet/outlet flux averaged
concentration to be a characteristic concentration produce unreliable results. In comparison, a few
unit cells open boundaries conditions resemble physical situation most and are harder to conduct
but provide the volumetric mass transfer coefficient based only on the spatial information for the
capillary number $Ca>0.7$. As well, a few unit cells simulations can provide more information about
mixing in the liquid slug in comparision with the periodic boundary conditions. We show
that all presented in literature strategies are extreme limits of one equation,
Eq. \ref{main:main:main}. Finally, the correlation for different Peclet numbers
is compared with analytical predictions presented by \citet{vanbaten-circular} and shows a good
comparison. The work is of use to people performing mass transfer numerical simulations for bubble
train flow in microchannels and within the
lattice Boltzmann context.
\end{abstract}

\section{Introduction}
\label{intro}
The monolith reactors during last years are getting more attention as alternative to slurry
reactors and trickle bed reactors \cite{kreutzer-overview,bercic-mass}. The bubble flow we focus in
this work, so-called Bretherton-Taylor flow \cite{bretherton,taylor}, describes the flow through
a liquid medium of equal sized, long in comparison with the channel height bubbles, see
Fig. \ref{fig:benchmark:hydro}. This flow regime
is usually characterized by surface tension dominance over inertia and viscous effects and by
comparatively small gas flow velocities \cite{yue-mass}. Due to surface tension dominance and small
scales of microchannels, bubble train flow exhibit advantageous properties which can not be achieved
in macroscopic counterparts: such as thin films \cite{bretherton} can remarkably enhance mass
transfer in the film from a gas or from walls; the plug flow nature that separates liquid slugs from each
other allowing to perform chemical reactions in slugs \cite{kreutzer-overview} or be an indicator of
the liquid velocity \cite{taylor}; the flow pattern depending on the capillary number $Ca$
\cite{heil-bretherton,heil-threedim}.  Overall, nowadays one can find a large number of applications
of Bretherton-Taylor bubble train flow: continuous flow analyzers, lung openings, chemical reactions
as hydrogenation of nitroaromatics, 2-ethyl-hexenal, Fischer-Tropsch synthesis, etc. One can find a
lot of different application references in extensive reviews \cite{kreutzer-overview,
gupta-review,yue-mass}.

@@

There are different methods to estimate the mass transfer coefficient $k_L$. We first examine the
theoretical definitions of the mass transfer in case of point mass sources.
\subsection{Point mass sources}
In what follows we will present three approaches to calculate point mass transfer
coefficients:
\begin{enumerate}
\item
Let us look at the infinitesimal small domain of the volume $A \Delta x$ not
moving and with the point mass source.  Then the concentration difference can be found as $\Delta C =
\cstar -
C(t)$, where $\cstar$ is the imposed point source concentration, $C(t)$ is
the time dependent concentration, which do not depend on the location due to homogeneity. Therefore, one can write the time dependent ODE for the
concentration in the domain:
\beq
\dot{m}= A \Delta x \frac{\mathrm{d}C}{\mathrm{d} t} = k_L P (\cstar-C(t)), 
\feq
with the initial condition $C(0)=0$
The solution can be found by solving ODE:
\beq
C(t)= \cstar (1-\exp(-\vol t )), 
\feq
where $\vol$ is the volumetric mass transfer coefficient defined as:
\beq
\vol=k_L \frac{P}{V},
\feq
where $P$ is the source surface, $V$ is the unit cell volume.
\item
Let us predict mass transfer happening in moving with the velocity
$U$ liquid, see Fig. \ref{fig:moving:frame}. 
\begin{figure}[htb!]
\includegraphics[width=\textwidth]{Figures/mass_transfer.eps}
\caption{The mass transfer in the moving liquid. \label{fig:moving:frame}}
\end{figure}

If one can assume that the mass point sources are distributed in the whole medium, then the mass accumulated in the unit volume $V=A \Delta x$ can be calculated as the difference of 
mass fluxes entering and leaving domain $U \bigl(C(x+\Delta x)-C(x)\bigr)$. We assume that the
source is a point source with the effective perimeter (area) $P$. The accumulated mass should be proportional
to the mass transfer coefficient:
\beq
U \bigl(C(x+\Delta x)-C(x)\bigr)=k_L P \bigl(\cstar-C(x)\bigr), 
\feq 
giving the same equation but only in the coordinate domain:
\beq
%\label{concentration:domain:coordinate}
C(x)= \cstar \Bigl(1-\exp\bigl(-k_L a \frac{x}{U} \bigr)\Bigr).
\label{main:mass:transfer:expression} 
\feq
Note that the concentration $C(x)$ does not depend on time.

@@

\subsection{Bubble train}
In the application to the bubble train flow, it is useful to think of one bubble to be a point
source to make all considerations above valid. For example, the expression
(\ref{main:mass:transfer:expression}) was used in the experiments by
\citet{bercic-mass}. However, one should be accurate with the definition of velocities because two
different phases co-exist in the bubble train flow. Usually, one can take the velocity $U$ to be as
a bulk velocity or $U=\ugas+\uliq$, where $\ugas$ and $\uliq$ are liquid and gas
superficial velocities correspondingly.

Thus, the experimental measurements are straightforward as measuring concentrations at
different
locations and calculating the volumetric mass transfer coefficient through the logarithmic
function. However, if one wants to analytically estimate the mass transfer coefficients, the
situation is much more complicated because of two phases presence and complicated bubble
geometry. Phases can have different mixing patterns
from film to slug and from slug to film. Also, as it was mentioned before depending on the capillary
number the mixing velocity pattern is different. Analytical approaches
\cite{irandoust,vanbaten-circular} assume that the
contributions from film and bubble caps can be calculated separately. However, one can see that
such 
an assumption (slug well mixed and the concentration is uniformly distributed over
the slug) usually overpredicts the mass transfer \cite{irandoust}. This happens since some tracer
concentration from film is mixed with the slug and changes the overall concentration in the slug.
Thus, all estimations for analytical mass trass calculation  do not account for mutual mass
transfer from
neighbouring bubbles.

Another issue is that if the slug is long enough then the film saturates with the tracer
concentration and its influence on mass transfer can be negligible \cite{vanbaten-circular}.
Mixing patterns of the film and liquid slugs are of great importance for the analytical
estimation of the mass transfer \cite{yue-mass}. However, the assumptions usually taken for the
mass transfer calculation are small capillary number and certain mixing patterns which help to
estimate the mass transfer using the penetration theory of \citet{higbie}.

In comparison with analytical calculations and simplifications, the numerical approach can take into
the account the complicated mixing patterns and geometries. However, there are challenges as to
mimic the continuous picture like moving with bulk velocity $U$ reacting medium as it is seen in
experiments. Thus, the questions indicated in Section \ref{intro}, as the choice of the
characteristic concentration, the choice of boundary conditions, arise. Next
section gives more details about numerical simulations.

\subsection{Numerical simulations}
\label{section:cases}
Ideally one needs to mimic the continuous picture as it is seen in experiments. Thus, one needs to
perform mass transfer simulations for a number of unit cells. As it was indicated above there are
two approaches towards it - either to simulate
the bubble train and then to measure concentration along the pipe, Eq.
\ref{main:mass:transfer:expression}, or to transfer to the reference frame moving with the bulk
velocity $U$ and conduct same measurements. However, both methods do require a tracking of
moving bubbles which is complicated from the numerical point of view. Therefore, one needs to come
up with simple smaller domain calculations of the mass transfer coefficient, which mimic the
continuous picture with infinite numbers of the bubbles closely.

@@

We perform simulations in the frame moving with the bubble (velocity
$\ububble$), that the bubble position stands steady. The bubble velocity $\ububble$ is
different from the bulk velocity $\ugas+\uliq$. In this case one needs to perform a coordinate $x$
variable change:
\beqal
\label{theor:average:concentration:time}
&x(t)=\ububble t\\
&C(x)=\cstar \Bigl(1-\exp\bigl(-\vol \frac{x}{\ugas+\uliq}\bigr)\Bigr)\\
&\langle C(t)\rangle=\cstar \Bigl(1-\exp\bigl(-\vol\, t \frac{\ububble}{\ugas+\uliq}\bigr)\Bigr),
\feqal
where $\langle C(t)\rangle$ is the characteristic concentration averaged in space
concentration. The geometry is not one-dimensional and one needs the characteristic concentration
to be only as a function of time. One can have different choices of
$\langle C(t) \rangle$ as the averaged in whole domain concentration or inlet/outlet averaged in
space concentrations.  Note that $C(x)$ is not averaged in time as $\overline{C(x)}$ since $C(x)$
ideally should not depend on time. 
Therefore, one can obtain the volumetric mass transfer coefficient through the concentration:
\beqal
\label{theor:one:concentration:time}
&\vol\, t \frac{\ububble}{\ugas+\uliq}=\ln \frac{\cstar}{\cstar-\langle C(t)\rangle}\\
&\volnondim=\frac{\lunit}{\ububble t}\ln \frac{\cstar}{\cstar-\langle C(t) \rangle},
\feqal
where the parameter $\vol \frac{\lunit}{\ugas+\uliq}$ is non-dimensional. One can also measure the
volumetric mass transfer coefficient from concentrations given at times $t_1$ and $t_2$:
\beq
\label{theor:continuous:mass:transfer}
\volnondim=\frac{\lunit}{\ububble
(t_2-t_1)}\ln\frac{C^{*}-\langle C(t_1) \rangle}{C^{*}-\langle C(t_2) \rangle}.
\feq
Expressions (\ref{theor:average:concentration:time}) and (\ref{theor:continuous:mass:transfer}) are
cornerstones of the present work. The main questions arisen in the present work are to define the
characteristic concentration $\langle C(t) \rangle$ and  boundary conditions to calculate a proper
volumetric mass transfer coefficient
$\vol$.

Three possible scenarios of conducting numerical simulations are examined in this work: 
\begin{enumerate}
\item The bubble train multiphase simulations are conducted by applying periodic boundary
conditions \cite{kuzmin-binary2d,kuzmin-binary3d}. One
can make an assumption of same boundary
conditions applied to the tracer concentration to simulate mass transfer. Therefore, only one unit
cell is simulated with
periodic boundary conditions, see Fig. \ref{fig:benchmark}. As well, no tracer leaves a
domain. The situation resembles a situation for the plug flow, where no tracer can leave a unit
domain. We expect such a situation resembles most the mass transfer for small Capillary numbers
$Ca<0.1$, where the liquid slug is well mixed and one have a homogenuous concentration in the unit
cell. Though easier to implement, it rises certain critisism
about the inlet concentration to be equal to the outlet concentration. In the continuous
formulation as it is seen in experiments there is the concentration difference between inlet and
outlet involving the volumetric
mass transfer coefficient, see second equation in system \ref{theor:one:concentration:time}.

One needs also address the choice of the characteristic concentration. For example, the
characteristic concentration $\langle C(t) \rangle$ required for calculation the volumetric mass
transfer coefficient, Eq.
\ref{theor:one:concentration:time}, can be taken as the average concentration in the domain:
\beq
\label{average:concentration:definition}
C(t)=\frac{\int_{liquid}{C \mathrm{d}V}}{\int{\mathrm{d}V}}.
\feq
\begin{figure}[htb!]
\includegraphics[width=\textwidth]{Figures/benchmark_middle.eps}\\
\caption{The two-dimensional benchmarks for the  the mass transfer
coefficient (bottom) for the bubble located near the entrance (top) and at the middle of the domain
(bottom). \label{fig:benchmark}}
\end{figure}
\item The second approach is the same as the first one but the characteristic concentration is taken
as in
\cite{vanbaten-circular}. The characteristic concentration is taken as the inlet/outlet
concentration. For the two-dimensional flow between plates inlet/outlet concentrations can be found
as:
\beqal
&\langle \cinlet(t) \rangle=\frac{\int{U(y) C(0,y,t) \mathrm{d}y}}{\int{U(0,y) \mathrm{d}y}}\\
&\langle \coutlet(t) \rangle=\frac{\int{U(y) C(\lunit,y,t) \mathrm{d}y}}{\int{U(\lunit,y)
\mathrm{d}y}}\\
&\cinlet(\bm{x},t)=\coutlet(\bm{x},t),\,\text{due to periodicity}.
\feqal
Therefore, the assumptions  of this approach is
that the difference between inlet and outlet is not
considerably large and the tracer is well mixed in the slug, so that the inlet or outlet
concentrations equal to the average concentration.

\item
The approach of \citet{vanbaten-circular} is examined. Periodic boundary conditions were used. We
used the
definition of the mass transfer coefficient to be proportional to the gain of the mass in the system
divided by the concentration difference
multiplied on the surface area:
\beq
\vol=\frac{\dot{m}}{P \Delta C}\frac{P}{V}=\frac{\dot{m}}{V (\cstar-\langle C(t) \rangle)},
\feq
where mass difference in the domain can be calculated as:
\beq
\dot{m}=\frac{m_2-m_1}{t_2-t_1}=\frac{\int_{liq}{C(\bm{x},t_2)\mathrm{d}V}-\int_{liq}{
C(\bm{x},t_1)\mathrm { d } V} } {
t_2-t_1 },
\feq
where the integration element $\mathrm{d}V=\mathrm{d}x\mathrm{d}y=\mathrm{d}\bm{x}$.
In the approach of \citet{vanbaten-circular} the inlet and outlet concentrations were taken as the
characteristic concentration $\langle C(t) \rangle $.

%\item 
%The approach of the opened boundaries was examined, see Fig. \ref{fig:benchmark:jos}. At the inlet
%and outlet boundaries the diffusion flux was imposed to be zero, $\frac{\partial C}{\partial x}=0$.
%The characteristic concentration was taken as the average concentration in the whole domain.
%\begin{figure}
%\includegraphics[width=\textwidth]{Figures/benchmark_jos.eps}
%\caption{The numerical mass transfer benchmark with open boundaries. The mass transfer is
%calculated according to Eq. \ref{theor:average:concentration:time} with the characteristic
%concentration accoring to Eq. \ref{average:concentration:definition}. \label{fig:benchmark:jos}}
%\end{figure}
\item 
The last approach to be examined in this paper is the approach of the simulation of a lot of unit
volumes, see Fig. \ref{fig:benchmark:alot}. This situation corresponds to the simulation of the
bubble train head, after injecting to the pipe and travelling along the pipe. One can see that this
situation corresponds most to the experimental picture. By simulation of a certain number of the
bubble train head, the influence of the boundaries can be
diminished. In this case, there is no ambiguity in the choice of the
characteristic concentration in this case. The average concentration in time will be governed
by Eq.
\ref{theor:continuous:mass:transfer}. This simulation statement is of most correspondence to
experiments and should give right answer. However, one of the
disadvantages is to simulate a certain number of unit cells ($1$ to $10$ in this case), thus
increasing a computational
complexity.
\begin{figure}[htb!]
\includegraphics[width=\textwidth]{Figures/benchmark_alot.eps}\\
\caption{Benchmark for a lot of units. \label{fig:benchmark:alot}}
\end{figure}
\end{enumerate}

One can notice that all the examined cases are extreme cases of one equation. The choice of boundary
conditions to use  or how many unit cells to simulate are not of importance but one needs only to
understand a proper mass balance for medium, which states:
\beq
\label{main:main:main}
\vol=\frac{\dot{m}-\int{\coutlet(t) u(\lunit,y)\mathrm{d}y}+\int{\cinlet(t) u(0,y)\mathrm{d}y}}{V
\Delta C}, 
\feq
where $\Delta C=\cstar-\langle C(t) \rangle$ with $\langle C(t) \rangle$ to be the average
concentration in whole liquid domain. Eq. \ref{main:main:main} is nothing else but a proper mass
balance: whatever was generated by the bubble surface equals to the mass domain change minus
whatever left domain plus whatever entered domain. Periodic boundary condition is the extreme
limiting case of Eq. \ref{main:main:main} as $\int{\coutlet(t)
u(\lunit,y)\mathrm{d}y}=\int{\cinlet(t) u(0,y)\mathrm{d}y}$. Thus, the mass change in the domain is
only because of the generating bubble surface. In this case, to calculate the volumetric mass
transfer coefficient one needs to track not only spatial information to calculate the averaged over
domain characteristic concentration  $\langle C(t) \rangle$ but as well time information to
calculate the mass change in the domain. It will be shown later, that for $Ca>0.7$ where velocity
pattern is changed, one can use only spatial information to calculate $\int{\cinlet(t) u(0,y)
\mathrm{d}y}$ and $\int{\coutlet(t) u(\lunit,y)\mathrm{d}y}$ as far as mass change in the domain is
zero, i.e. $\dot{m}=0$.

Before we examine all the test cases above, the lattice Boltzmann mass transfer
benchmarks will be presented.

\section{Validation}
For analytical expressions the problem of the mass transfer can be decomposed to the following
parts: the mass transfer from two circumferences and the mass transfer from the film. We will
examine
these problems with the help of the lattice Boltzmann method and compare them against analytical
solutions. The next sections will present the lattice Boltzmann method and a few benchmarks.

\subsection{TRT $D2Q9$ model}
The lattice Boltzmann equation (LBE) operates on a square/cubic grid representing the
physical domain. It utilizes
probability distribution functions (also known as particle populations)
containing information about
macroscopic variables, such as fluid density and momentum. LBE consists of
two parts: a local collision step, and a propagation step which transports
information from one node to another along some 
directions specified by the discrete velocity set.
The LBE is typically implemented as follows:
\begin{equation}
\label{standard:implementation}
\begin{aligned}
&f_i^{*}(\bm{x},t)= f_i(\bm{x},t)-\omega \bigl(f_i(\bm{x},t)-eq_i(\bm{x},t)\bigr),&&\text{
collision step}\\
&f_i(\bm{x}+\bm{c_i},t+1)=f_i^{*}(\bm{x},t),&&\text{ propagation step}, 
\end{aligned}
\end{equation}
where $f_i$ is the probability distribution function in the direction $\bm{c_i}$,
 $f_i^{eq}$ is the equilibrium probability distribution function, $\omega$ is the
relaxation parameter. Collision operator, $-\omega \bigl(f_i- f_i^{eq}\bigr)$ is so-called BGK
collision operator \cite{bgk}. However, the approach to be used here is the TRT
(two-relaxation-times) collision operator \cite{ginzburg-main,ginzburg-saturated-flow}. In
comparison with the widely used BGK collision operator, TRT collision operator has better accuracy
for diffusion and convection fluxes, and larger range of parameters where the LBM scheme is stable.
